---
title: "Cause of Death"
output: html_notebook
---

# Introduction

The aim of this repo is to investigate the change in CVD and non-CVD mortality in England & Wales over time, in comparison with a couple of other countries 

The data source used will be the [Human Cause of Death](https://www.causesofdeath.org/) database. THe source it uses is the [WHO Mortality Database](https://www.who.int/healthinfo/statistics/mortality_rawdata/en/) 

The information required to decipher the data are available [here](https://www.causesofdeath.org/docs/formats.pdf)

For the short code:

* 0   All causes
* 7   Heart Diseases
* 8   Cerebrovascular diseases

```{r}
pacman::p_load(tidyverse, here, readxl)


```

Load SDR short list

```{r}

sdr <- read_csv(here("data", "GBRTENW", "GBRTENW_sdr_short_idr.csv"))

```


```{r}
short_code_lookup <- read_excel(here("data", "lookups.xlsx"), sheet = "cause_short")
sex_lookup <- read_excel(here("data", "lookups.xlsx"), sheet = "sex")
```

Join

```{r}
tidied_sdr <- 
  sdr %>% 
    left_join(sex_lookup, by = c("sex" = "no")) %>% 
    select(year, sex = sex.y, cause, sdr = SDR2013) %>% 
    left_join(short_code_lookup, by = c("cause"= "no")) %>% 
    select(year, sex, cause = short_title, sdr)

tidied_sdr

```


```{r}

tidied_sdr %>% 
  filter(cause %in% c("all")) %>% 
  ggplot(aes(x = year, y = sdr, group = cause, colour = cause)) + 
  geom_line() +
  facet_wrap(~sex)

tidied_sdr %>% 
  filter(cause %in% c("circulatory", "cerebrovascular")) %>% 
  ggplot(aes(x = year, y = sdr, group = cause, colour = cause)) + 
  geom_line() +
  facet_wrap(~sex)

tidied_sdr %>% 
  group_by(sex, cause) %>% 
  arrange(year) %>% 
  mutate(ch_sdr = sdr - lag(sdr)) %>% 
  ungroup() %>% 
  filter(cause %in% c("cerebrovascular")) %>% 
  ggplot(aes(x = year, y = ch_sdr, group = cause, colour = cause)) + 
  geom_point() +
  stat_smooth() +
  geom_hline(yintercept = 0) +
  facet_wrap(~sex)

tidied_sdr %>% 
  group_by(sex, cause) %>% 
  arrange(year) %>% 
  mutate(ch_sdr = sdr - lag(sdr)) %>% 
  ungroup() %>% 
  filter(cause %in% c("all")) %>% 
  ggplot(aes(x = year, y = ch_sdr, group = cause, colour = cause)) + 
  geom_point() +
  stat_smooth() +
  geom_hline(yintercept = 0) +
  facet_wrap(~sex)

```


It's difficult to tell with this, the period is too short. It might be better to go to the WHO database instead....